Many claims of group polarization are plausibly false due to inappropriate statistics

### 2024-01-19 GroupPolarizationStatmod/paper/ms.tex
updated figures and started metric model analysis

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{float}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{caption}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\begin{abstract}
  \noindent
  % Psychology is beset by three crises: the reproducibility crisis, the
  % generalizability crisis, and the theory crisis. New verbal theories for various
  % phenomena are constantly being developed orthogonal to other theories, all based on
  % mostly unreproducible analyses; even if a positive result is in some sense ``true''
  % it may not account for all the variance present in a set of observations---studies
  % thus fail to generalize because their results do not account for this variance.
  % There is another related problem: a phenomenon may be ``reproducible'' only
  % because studies of the phenomenon systematically misuse statistics in such a way
  % that induces false positives for each succeeding ``replication.'' 
  \emph{Group polarization} is a theoretical process where like-minded novel group members 
  become more extreme in their opinions after group discussion, but also find \emph{consensus}, reducing 
  group-level opinion variance.
  \emph{Simple consensus} may occur instead, where opinion variance decreases but
  the mean group opinion does not change.
  Here we show that many published detections of group polarization are explained
  equally rigorously as simple consensus.
  This is the case because experimenters failed
  to account for the measurement of continous-valued psychological opinions 
  as ordinal-valued reported opinions, e.g., on a Likert-type scale. 
  This creates ceiling effects that mask whether extreme opinions become less extreme.
  We developed a generative model to demonstrate that 95\% of sixty experimental conditions
  across ten group polarization papers report plausibly false detections of
  group polarization. Our approach can be used to evaluate whether other detections of
  group polarization may be plausibly false---our approach could be applied beyond
  group polarization as well. This work also provides a
  framework for new group polarization studies that could
  share a common, rigorous, and interoperable study design to multiply our knowledge
  of this potentially important group process, avoiding the historically fragmented
  approach using disparate, incommensurate study designs.

  % To show this, we developed a
  % novel generative model that provides two levels of evidence that an historical
  % published empirical finding of group polarization is plausibly false.  Our model
  % can be used to evaluate additional studies in group polarization and beyond, even
  % if the reported data and significant metrics are incompletely reported, and the
  % original raw data has been lost to time.  This may be just the tip of the
  % plausibly-false-positives-in-psychology iceberg, both for group polarization
  % studies and for the legion psychological results that model ordinal observations of
  % opinions, beliefs, emotions, or any other latent psychological variable that is
  % theoretically continuous, but measured using an ordinal scale.  
\end{abstract}
\clearpage

@@

\begin{table}[h]
  \caption{Variables used in the formal model of group polarization from simple consensus.}
  \label{tab:modelVariables}
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $\mu$  &   Latent mean, assumed constant for simple consensus \\ 
   $\sigma_t$ & Latent standard deviation at time $t$ \\
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma_t^2$ \\
   $k \in \{0,\ldots,K\} $ & Threshold index ($k$) between 0 and total number of
         ordinal bins, $K$ \\
   $b_k$ & Opinion bin $k$; note there is no $k=0$ bin. \\
   $\theta_{k-1},~\theta_k$  & Lower, upper threshold value for binning latent opinions into
     bin $b_k$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\langle o_t \rangle$  &   Observed mean opinion calculated from ordinal data at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

@@

Here we evaluate if metric linear models inferred group polarization from
simulated simple consensus data using parameters identified above. Those
parameters specified continous pre- and post-deliberation

% \spacing{1.0}
% \hspace{-1em}
% \input{t-test-table.tex}
% % \input{t-test-table-custom.tex}
% \spacing{1.5}

@@

@@

\section{Notes on studies}

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{float}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\begin{abstract}
  \noindent
  \emph{Group polarization} is a theoretical process where like-minded novel group members 
  become more extreme in their opinions after group discussion, but also find \emph{consensus}, reducing 
  group-level opinion variance.
  \emph{Simple consensus} may occur instead, where opinion variance decreases but
  the mean group opinion does not change.
  Here we show that many published detections of group polarization are explained
  equally rigorously as simple consensus.
  This is the case because experimenters failed
  to account for the measurement of continous-valued psychological opinions 
  as ordinal-valued reported opinions, e.g., on a Likert-type scale. 
  This creates a ceiling effect that masks extreme opinions becoming less extreme.
  We developed a generative model to 
  demonstrate that 95\% of sixty experimental conditions
  across ten group polarization papers report plausibly false detections of
  group polarization. Our approach can be used to evaluate whether other detections of
  group polarization may be plausibly false---our approach could be applied beyond
  group polarization as well. This work also provides a
  framework for new group polarization studies that could
  share a common, rigorous, and interoperable study design to multiply our knowledge
  of this potentially important group process, avoiding the historically fragmented
  approach using disparate, incommensurate study designs.
\end{abstract}

\clearpage

@@

\begin{table}[h]
  \caption{Variables used in the formal model of group polarization from simple consensus.}
  \label{tab:modelVariables}
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $\mu$  &   Latent mean, assumed constant for simple consensus \\ 
   $\sigma_t$ & Latent standard deviation at time $t$ \\
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma_t^2$ \\
   $k \in \{0,\ldots,B\} $ & Threshold index ($k$) between 0 and total number of
         ordinal bins, $B$ \\
   $b_k$ & Opinion bin $k$; note there is no $k=0$ bin. \\
   $\theta_{k-1},~\theta_k$  & Lower, upper threshold value for binning latent opinions into
     bin $b_k$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\langle o_t \rangle$  &   Observed mean opinion calculated from ordinal data at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

@@

Here we evaluate if metric linear models inferred group polarization from simulated
simple consensus data using parameters identified above. Those parameters specified
continous pre- and post-deliberation means and variance of normal opinion
distributions that represent simple consensus but yields
pre- and post-deliberation ordinal distributions with different, group-polarized
means. This parameter-finding process implicitly assumes $N\to\infty$ participants.

It is therefore necessary to evaluate whether the simple consensus parameters so 
identified in the previous step yield simulated data that also appear to be
group-polarized when measured on an ordinal scale. We draw $N$ samples from the
pre- and post-deliberation normal distributions and bin them into one of 
$B$ ordinal values. $N$ and $B$ are set specifically for each of the 
60 experimental conditions analyzed here. We then fit a linear model to the 
simulated observed pre- and post-deliberation data to infer whether the mean
shifted, indicating group polarization. We measure the pre- and
post-deliberation standard deviations as well. Then we calculate Cohen's $d$
to quantify the effect size for each simulation. A suite of simulations are
necessary since there might be significant variance in $d$ for different
generations of simulated pre- and post-deliberation observations. Therefore we
perform 1000 trials for each experimental condition and inspect the resulting
distributions of $d$ across trials in a boxplot.

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:MetricBoxplot_N} ABOUT HERE ]} \\
\end{center}

@@

We did make some simplifying assumptions that could affect model outcomes.

@@

\section*{Figures}

\begin{figure}
  \caption{
    Boxplots visualizing the distribution of Cohen's $d$ values (x-axis in
    A and B) inferred through a metric linear model over 1000 simulation 
    trials for each experimental condition (y-axis in A and B). All conditions
    have many trials that result in non-zero Cohen's $d$ values with number of
    simulated participants, $N$, set to their empirical values reported in the
    associated journal article (A). However, when the number of simulated
    participants is increased to $10N$, there are several conditions that 
    indeed yield distributions closer to $d = 0$ as expected, however some
    conditions have $d$ distributions centered on a nonzero mean with little to
    no overlap with $d=0$; variance in $d$ decreases as expected
    across experimental conditions (B).
  }  
  \label{fig:MetricBoxplot}
  \centering
  \begin{subfigure}[]{0.9\textwidth}
    % \hspace{-7.5em} 
    \centering
    \textsf{\large \textbf{A}, empirical $N$} \\
    \includegraphics[width=\textwidth]{Figures/boxplots/metric_cohens.pdf}
    \label{fig:MetricBoxplot_N}
  \end{subfigure} \\
\end{figure}
\begin{figure}\ContinuedFloat
  \begin{subfigure}[]{0.9\textwidth}
    % \hspace{-5.5em} 
    \centering
    \textsf{\large \textbf{B}, 10 $\times$ empirical $N$} \\
    \includegraphics[width=\textwidth]{Figures/boxplots/10N_metric_cohens.pdf}
    \label{fig:MetricBoxplot_10N}
  \end{subfigure}
\end{figure}

\clearpage

\section*{Notes on studies}
